---
title: "5110 project"
author: "Lesrene Browne"
date: "2023-04-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tibble)
library(dplyr)
library(dtplyr)
library(tidyr)
library(data.table)
library(naniar)
library(broom)
library(modelr)
library(stringr)
library(MASS)
library(ggplot2)
```

# Building dataset and tidying from Numerator File
```{r}
# Tidy Process
  # Read in the data
  # Arrange birthweight in desc order
  # Replace all 9999 values in birthweight and DOB Time with NAs and omit
  # Generate unique key from birthweight, DOB time, facility type, 
  #      pre-pregnancy weight, weight gain, DOB Year, DOB Month, DOB Week 
  # Remove duplicate keys 

numerator_data <- fread("/Users/lesrene/Downloads/link2020usnum.csv", 
                      select=c('brthwgt','dob_tt', 'bfacil','pwgt_r', 
                               'wtgain', 'dob_yy', 'dob_mm',
                               'dob_wk', 'sex')) %>% arrange(desc(brthwgt))
numerator_data <- data.frame(numerator_data)
numerator_data$brthwgt <- replace(numerator_data$brthwgt, 
                                  numerator_data$brthwgt == 9999, NA)
numerator_data$dob_tt <- replace(numerator_data$dob_tt, 
                                 numerator_data$dob_tt == 9999, NA)
numerator_data <- numerator_data %>% na.omit()
numerator_data <- numerator_data %>% mutate(key = paste(numerator_data$brthwgt,
                                            numerator_data$dob_tt,
                                            numerator_data$bfacil,
                                            numerator_data$pwgt_r,
                                            numerator_data$wtgain,
                                            numerator_data$dob_yy,
                                            numerator_data$dob_mm,
                                            numerator_data$dob_wk,
                                            sep = '-'))
numerator_data <- distinct(numerator_data)
```

# Building dataset and tidying from Denominator File
# Created mortality_status indicator column (0 and 1)
```{r}
# Build dataset from Denominator File
denom_data <- fread("/Users/lesrene/Downloads/link2020usden.csv",
                    select=c('brthwgt','dob_tt','bfacil','pwgt_r','wtgain',
                             'mrace6','ilive','meduc','bfacil3','rf_pdiab',
                             'rf_gdiab','rf_phype','rf_ghype','rf_ehype',
                             'rf_ppb','rf_inft','rf_drg','rf_art','rf_cesar',
                             'no_risks', 'mager9', 'dob_yy', 'dob_mm','dob_wk',
                             'sex'))
# Tidying data
denom_data <- data.frame(denom_data)
denom_data <- denom_data %>% arrange(desc(brthwgt)) 
denom_data$brthwgt <- replace(denom_data$brthwgt, 
                              denom_data$brthwgt == 9999, NA)
denom_data$dob_tt <- replace(denom_data$dob_tt, 
                             denom_data$dob_tt == 9999, NA) 
denom_data <- denom_data %>% na.omit()
denom_data <- denom_data %>% mutate(key = paste(denom_data$brthwgt,
                                            denom_data$dob_tt,
                                            denom_data$bfacil,
                                            denom_data$pwgt_r,
                                            denom_data$wtgain,
                                            denom_data$dob_yy,
                                            denom_data$dob_mm,
                                            denom_data$dob_wk,
                                            sep = '-'))

denom_data <- distinct(denom_data)
# Creating the mortality indicator column 
# create a new column in the births dataframe to indicate mortality status
denom_data$mortality_status <- 0
# join the two dataframes based on the primary key column
merged_df <- left_join(denom_data, numerator_data, by = "key")
# update the mortality status column based on whether or not a match was found
merged_df$mortality_status[!is.na(merged_df$brthwgt.y)] <- 1
```


# Sampling unbalanced Deaths vs Births

#The first line of code death_cases <- merged_df[merged_df$death == 1, ] is selecting all rows from merged_df where the death column has a value of 1 and assigning those rows to a new data frame named death_cases.

#The second line of code alive_cases <- merged_df[merged_df$death == 0, ] is selecting all rows from merged_df where the death column has a value of 0 and assigning those rows to a new data frame named alive_cases.

#So, death_cases will contain all the rows from merged_df where someone has died, while alive_cases will contain all the rows where someone is still alive. This could be useful for analyzing the factors that contributed to death or for comparing the characteristics of those who survived versus those who did not.
```{r}
# Subset the data frame by death and alive classes
death_cases <- merged_df[merged_df$mortality_status == 1, ]
alive_cases <- merged_df[merged_df$mortality_status == 0, ]
# Compute the number of cases in the minority class
n_death <- nrow(death_cases)
# Sample n_alive cases from the majority class 
n_alive <- n_death 
alive_sample <- alive_cases[sample(nrow(alive_cases), n_alive), ]
# Combine the resampled data frames into a new data frame
balanced_data <- rbind(death_cases, alive_sample)
head(balanced_data)
```

# The code above is performing the task of balancing the data frame by oversampling the minority class (i.e., death_cases) using a random sample of the majority class (i.e., alive_cases). This is a common technique in machine learning when the two classes are imbalanced, meaning that one class has significantly fewer observations than the other.


# Step 1: Calculate the risk level based on columns 313-331
# Step 2: Create 3 categories based on risk level number: no/low risk, medium risk, high risk (10 risk factors: 0-2, 3-5, 6-10)
# Step 3: Use this variable to see if it will make the model better fit 

```{r}

# Removing columns that created the key
births <- balanced_data %>% dplyr::select(-c('brthwgt.x','brthwgt.y', 'dob_tt.x','dob_tt.y',
                                                  'bfacil.x','bfacil.y', 
                                                 'pwgt_r.x', 'pwgt_r.y','wtgain.x', 
                                                 'wtgain.y'))
# Next steps: 
  # Convert the Y/N in risk columns to 0 and 1's
  # Sum up all the fisk factor columns for risk score
  # Create risk level column and create category (Low, Medium, High) based on score
imputed_rf <- births %>% 
  mutate_at("ilive", ~str_replace(.,"N", "0")) %>%
  mutate_at("ilive", ~str_replace(.,"Y", "1")) %>%
  mutate_at(vars(contains("rf_")), ~str_replace(.,"N", "0")) %>%
  mutate_at(vars(contains("rf_")), ~str_replace(.,"Y", "1")) %>%
  mutate_at(vars(contains("rf_")), ~str_replace(.,"X", "0")) %>%
  mutate_at(vars(contains("rf_")), ~str_replace(.,"U", "0")) %>%
  mutate_at(vars(contains("rf_")), ~as.numeric(as.character(.))) %>%
  mutate(sumrf = rowSums(dplyr::select(., contains("rf_")))) %>%
  mutate(risk_level_cat = ifelse(sumrf %in% 0:5, 1, ifelse(sumrf %in% 6:11, 2,
                                                           ifelse(sumrf %in% 12:17, 3, NA)))) %>%
  mutate(risk_level = ifelse(sumrf %in% 0:2, "Low", ifelse(sumrf %in% 3:5, "Medium",
                      ifelse(sumrf %in% 6:10, "High", NA))))
# Data Clean up to remove some columns that are not needed
tibble(imputed_rf)
```

# Are we still keeping this in here? 
```{r}
# fit the logistic regression model
model_simple <- glm(mortality_status ~ mrace6, data = imputed_rf, family = binomial)
summary(model_simple)
```




## Partitioning the dataset
```{r}
set.seed(2)
imputed_partition <- resample_partition(imputed_rf,
                                    p=c(train=0.6,
                                        valid=0.2,
                                        test=0.2))
imputed_partition
```


# Stepwise selection using AIC
```{r}

# Use step AIC to select the best number of variables to include in the model based on AIC
model_stepwise_AIC <- stepAIC(glm(mortality_status ~ 1, 
                                  data = imputed_partition$train, 
                                  family = binomial(link = "logit")), 
                      direction = "forward", 
                      scope = formula(~ mrace6 + risk_level_cat + 
                                        meduc + mager9), 
                      trace = FALSE, 
                      k = log(nrow(imputed_partition$train)))
# Print the selected model
summary(model_stepwise_AIC)
# Test for the effect of race on infant mortality using odds ratios
#exp(cbind(OR = coef(model_stepwise_AIC), confint(model_stepwise_AIC)))
```


-------------------------------------------------------------------------------
# Stretch goal work


```{r}
# fit the logistic regression model
model <- glm(mortality_status ~ mrace6 + mager9 + meduc, data = imputed_rf, family = binomial)
summary(model)
```


```{r}
# may end up needing to weight data  by race after all but luckily there's infant mortality rates documented by race in the documentation maybe we can use those to weight 
# The infant mortality rates based on the documentation for maternal race:
# Overall = 5.51
# Non-Hispanic White = 4.45
# Non-Hispanic Black = 10.6
# Non-Hispanic American Indian or Alaska Native = 7.63
# Non-Hispanic Asian = 3.25
# Non-Hispanic Native Hawaiian or Other Pacific Islander = 7.68
# Hispanic = 4.87
```

```{r}
data_race <- imputed_rf
data_race$mrace6 <- 1
fitted_probs <- predict(model, newdata = data_race, type = "response")
sim_outcome_1 <- rbinom(n = nrow(data_race), size = 1, prob = fitted_probs)
data_race$mrace6 <- 2
fitted_probs <- predict(model, newdata = data_race, type = "response")
sim_outcome_2 <- rbinom(n = nrow(data_race), size = 1, prob = fitted_probs)
# perform t-test
p_val <- t.test(sim_outcome_2, sim_outcome_1)$p.value
p_val
```

```{r}
# set up the range of values to vary
#race_values <- unique(imputed_rf$mrace6)
race_values <- seq(from = 1, to = 6, by = 1)
# create function to simulate outcome data based on varying maternal race values
# got this function w/ help from chatgpt
sim_outcome <- function(data, maternal_race) {
  data$maternal_race <- maternal_race
  fitted_probs <- predict(model, newdata = data, type = "response")
  sim_outcome <- rbinom(n = nrow(data), size = 1, prob = fitted_probs)
  return(sim_outcome)
}
# simulate outcome data for varying maternal race values
sim_data <- lapply(race_values, function(mr) sim_outcome(data_race, mr))
sim_data <- do.call(cbind, sim_data)
# calculate odds ratios for each simulated dataset
odds_ratios <- apply(sim_data, 2, function(x) {
  model_sim <- glm(x ~ mrace6 + mager9 + meduc, data = imputed_rf, family = binomial)
  exp(coef(model_sim)["mrace6"])
})
odds_ratios
```

```{r}
# plot odds ratios as a function of maternal race
plot(race_values, odds_ratios, type = "l", xlab = "Maternal Race", ylab = "Odds Ratio")
```


```{r}
# create empty vectors to store odds ratios and p-values for each race group
p_values <- c(0, 0, 0, 0, 0, 0)
# loop through each maternal race group
for(i in 1:length(race_values)) {
  
  # calculate p-value for current race group
  model_race <- glm(mortality_status ~ mrace6 + mager9 + meduc, data = data_subset, family = binomial)
  p_value_i <- summary(model_race)$coef[2,4]
  
  # store results for current race group in vectors
  odds_ratios[i] <- median(odds_ratios_i)
  p_values[i] <- p_value_i
}
```


```{r}
p_values
# plot p-values for each maternal race group
barplot(p_values, names.arg = race_values, ylab = "p-value")
abline(h = 0.05, lty = 2)
```



```{r}
# Visualization 
tibble(numerator_data)
tibble(denom_data)
#numerator_data
numerator_counts <- numerator_data %>%
  select(mrace6) %>%
  count(mrace6) %>%
  rename(num_counts = n) %>%
  group_by(mrace6)
numerator_counts <- data.frame(numerator_counts)
denom_counts <- denom_data %>%
  select(mrace6) %>%
  count(mrace6) %>%
  rename(denom_counts = n) %>%
  group_by(mrace6)
denom_counts <- data.frame(denom_counts)
combined_data <- cbind.data.frame(numerator_counts, denom_counts$denom_counts)%>%
  mutate(proportion = (num_counts/denom_counts$denom_counts)*1000)
combined_data
```



```{r}

ggplot(data, aes(fill=mrace6, y=proportion, x=mrace6)) + 
    geom_bar(stat = "identity")+ theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(title="Race Distribution for Trans People (Weighted & Unweighted)")
```

Does Race affect Infant Mortality Rate ?

- H0: there is no relationship between Race and IMR

- H1: there is a relationship between Race and IMR

```{r}
datatest <- data1 %>%
  #filter(cut %in% c("Fair", "Ideal")) %>%
  transmute(co_seqnum, co_dodyy)
datatest <- na.omit(datatest)
datatest
data2$co_dodyy
diamonds2class
typeof(data1$co_seqnum)
typeof(data2$co_seqnum)
              
g <- inner_join()
g
```