---
title: "5110 project"
author: "Lesrene Browne"
date: "2023-04-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tibble)
library(dplyr)
library(dtplyr)
library(tidyr)
library(data.table)
library(naniar)
library(broom)
library(modelr)
library(stringr)
library(MASS)
library(ggplot2)
library(caret)
library(viridis)
```

# --------------------------  DATA PREP  ---------------------------------------

## Building dataset and tidying from Numerator File
```{r}
# Tidy Process
  # Read in the data
  # Arrange birthweight in desc order
  # Replace all 9999 values in birthweight and DOB Time with NAs and omit
  # Generate unique key from birthweight, DOB time, facility type, 
  #      pre-pregnancy weight, weight gain, DOB Year, DOB Month, DOB Week 
  # Remove duplicate keys 
# Path files for running locally
#/Users/lesrene/Downloads/link2020usnum.csv"
# 
numerator_data <- fread("./link2020usnum.csv",
                      select=c('brthwgt','dob_tt', 'bfacil','pwgt_r',
                               'wtgain', 'dob_yy', 'dob_mm',
                               'dob_wk', 'sex')) %>% arrange(desc(brthwgt))
numerator_data <- data.frame(numerator_data)
numerator_data$brthwgt <- replace(numerator_data$brthwgt, 
                                  numerator_data$brthwgt == 9999, NA)
numerator_data$dob_tt <- replace(numerator_data$dob_tt, 
                                 numerator_data$dob_tt == 9999, NA)
numerator_data <- numerator_data %>% na.omit()
numerator_data <- numerator_data %>% mutate(key = paste(numerator_data$brthwgt,
                                            numerator_data$dob_tt,
                                            numerator_data$bfacil,
                                            numerator_data$pwgt_r,
                                            numerator_data$wtgain,
                                            numerator_data$dob_yy,
                                            numerator_data$dob_mm,
                                            numerator_data$dob_wk,
                                            numerator_data$sex,
                                            sep = '-'))
numerator_data <- distinct(numerator_data)
```

# Building dataset and tidying from Denominator File
# Created mortality_status indicator column (0 and 1)
```{r}
# Build dataset from Denominator File
#/Users/lesrene/Downloads/link2020usden.csv
denom_data <- fread("./link2020usden.csv",
                    select=c('brthwgt','dob_tt','bfacil','pwgt_r','wtgain',
                             'mrace6','ilive','meduc','bfacil3','rf_pdiab',
                             'rf_gdiab','rf_phype','rf_ghype','rf_ehype',
                             'rf_ppb','rf_inft','rf_drg','rf_art','rf_cesar',
                             'no_risks', 'mager9', 'dob_yy', 'dob_mm','dob_wk',
                             'sex'))
# Tidying data
denom_data <- data.frame(denom_data)
denom_data <- denom_data %>% arrange(desc(brthwgt)) 
denom_data$brthwgt <- replace(denom_data$brthwgt, 
                              denom_data$brthwgt == 9999, NA)
denom_data$dob_tt <- replace(denom_data$dob_tt, 
                             denom_data$dob_tt == 9999, NA) 
denom_data <- denom_data %>% na.omit()
denom_data <- denom_data %>% mutate(key = paste(denom_data$brthwgt,
                                            denom_data$dob_tt,
                                            denom_data$bfacil,
                                            denom_data$pwgt_r,
                                            denom_data$wtgain,
                                            denom_data$dob_yy,
                                            denom_data$dob_mm,
                                            denom_data$dob_wk,
                                            denom_data$sex,
                                            sep = '-'))
denom_data <- distinct(denom_data)
# Creating the mortality indicator column 
# create a new column in the births dataframe to indicate mortality status
denom_data$mortality_status <- 0
# join the two dataframes based on the primary key column
merged_df <- left_join(denom_data, numerator_data, by = "key")
# update the mortality status column based on whether or not a match was found
merged_df$mortality_status[!is.na(merged_df$brthwgt.y)] <- 1
```


```{r}
# Removing columns that created the key
births <- merged_df %>% dplyr::select(-ends_with(".x"), -ends_with(".y"))
# Tidying:
# Convert the Y/N in risk columns to 0 and 1's
# Sum up all the risk factor columns to calculate maternal health risk score
# Create risk level column and assign category
#(No/Low, Medium, High) based on score
full_data_clean <- births %>%
 mutate_at("ilive", ~str_replace(.,"N", "0")) %>%
 mutate_at("ilive", ~str_replace(.,"Y", "1")) %>%
 mutate_at(vars(contains("rf_")), ~str_replace(.,"N", "0")) %>%
 mutate_at(vars(contains("rf_")), ~str_replace(.,"Y", "1")) %>%
 mutate_at(vars(contains("rf_")), ~str_replace(.,"X", "0")) %>%
 mutate_at(vars(contains("rf_")), ~str_replace(.,"U", "0")) %>%
 mutate_at(vars(contains("rf_")), ~as.numeric(as.character(.))) %>%
 mutate(sumrf = rowSums(dplyr::select(., contains("rf_")))) %>%
 mutate(risk_level_cat = ifelse(sumrf %in% 0:2, 1,
                                ifelse(sumrf %in% 3:5, 2, 
                                       ifelse(sumrf %in% 6:10, 3, NA)))) %>%
 mutate(risk_level = ifelse(sumrf %in% 0:2, "Low", 
                            ifelse(sumrf %in% 3:5, "Medium",
                                   ifelse(sumrf %in% 6:10, "High", NA))))
  

# Data Clean up to remove some columns that are not needed
full_data_clean <- full_data_clean %>% dplyr::select(-starts_with("rf_"))

tibble(full_data_clean)
```

Adding Categorical Columns Associated with our variables 
```{r}
# Age Lookup table
age_table <- data.frame(
  age_category = c("< 15", "15-19", "20-24", "25-29", "30-34", "35-39", "40-44", "45-49", "50-54"),
  numerical_value = c(1, 2, 3, 4, 5, 6, 7, 8, 9)
)
# Race Lookup Table
race_table <- data.frame(
  race_category = c("White", "Black", "American Indian/Alaskan", "Asian", "Native Hawaiian", "More than Race"),
  numerical_value = c(1, 2, 3, 4, 5, 6)
)
# Education Lookup Table
education_table <- data.frame(
  education_category = c("(1) 8th Grade or Less", "(2) 9-12th Grade w/out Diploma", "(3) High School Graduate", "(4) Some College Credits", "(5) Associate's", "(6) Bachelor's",
                         "(7) Master's", "(8) Doctorate/Professional Degree","(9) Unknown"),
  numerical_value = c(1, 2, 3, 4, 5, 6,7,8,9)
)

# Facility Lookup Table
facility_table <- data.frame(
  facility_category = c("(1) In Hospital", "(2) Not in Hospital", "(3) Unknown or Not Stated"),
  numerical_value = c(1, 2, 3)
)

full_data_clean <- full_data_clean %>% 
  mutate(Age_Category = age_table$age_category[match(full_data_clean$mager9, 
                                                  age_table$numerical_value)]) %>%
    mutate(Race_Category = race_table$race_category[match(full_data_clean$mrace6, 
                                                  race_table$numerical_value)]) %>%
      mutate(Educ_Category = education_table$education_category[match(full_data_clean$meduc, 
                                                  education_table$numerical_value)]) %>%
      mutate(Facil_Category = facility_table$facility_category[match(full_data_clean$bfacil3, 
                                                  facility_table$numerical_value)])
```


# --------------------------  EDA  ---------------------------------------------

### Distributions
# Distribution of Race
```{r}
ggplot(full_data_clean, aes(x = Race_Category, fill = Race_Category)) + 
  geom_bar() + 
  scale_fill_discrete(name = "Race") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(title = "Distribution of Race", x = "Race", y = "Count")
```

# Distribution of Age 
```{r}
ggplot(full_data_clean, aes(x=Age_Category, fill = Age_Category)) + 
  geom_bar()+ 
  scale_fill_discrete(name = "Age Ranges") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(title="Distribution of Age", x = "Age Ranges", y = "Count")
```

# Distribution of Health Risk Level
```{r}
ggplot(full_data_clean, aes(x=risk_level, fill = risk_level)) + 
  geom_bar()+ 
  scale_fill_discrete(name = "Risk Level") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(title="Distribution of Health Risk Level", x = "Risk Level", y = "Count")
```

# Distribution of Education
```{r}
ggplot(full_data_clean, aes(x=Educ_Category, fill = Educ_Category)) + 
  geom_bar()+ 
  scale_fill_discrete(name = "Education Level") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(title="Distribution of Maternal Education Level", x = "Education", y = "Count")
```

# Distribution of Birth Facility
```{r}
ggplot(full_data_clean, aes(x=Facil_Category, fill = Facil_Category)) + 
  geom_bar()+ 
  scale_fill_discrete(name = "Facility of Birth") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(title="Distribution of Birth Facility", x = "Facility/Location", y = "Count")
```

### Evaluating Relationships between variables

# Relationship between Age and Race
```{r}
ggplot(full_data_clean, aes(x=Age_Category, fill=Race_Category)) + 
  geom_bar(position = "stack")+ 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(title="Relationship between Age and Race") 
  #facet_grid(~ Race_Category)
```

# Relationship between Age and Risk Level HeatMap
```{r}
freq_table <- table(Age = full_data_clean$Age_Category, Risk_Level = full_data_clean$risk_level_cat)

ggplot(as.data.frame(freq_table), aes(x = Age, y = Risk_Level, fill = Freq)) +
  geom_tile() +
  scale_fill_viridis(name = "Count") +
  labs(title = "Relationship Between Age and Health Risk Level",
       x = "Age", y = "Risk Level")

```



# ------------------------  DATA PARTIITIONING ---------------------------------

# Partitioning the data (training, testing)
```{r}
set.seed(2)
imputed_partition <- resample_partition(full_data_clean,
                                      c(train=0.6,
                                        test=0.4))
```


## PART A
# --------------------------  STATISTICAL INFERENCE  ---------------------------

# Generating model from stepAIC model
```{r}
# Use step AIC to select the best number of variables to include in the model based on AIC
model_stepwise_AIC <- stepAIC(glm(mortality_status ~ 1, data = imputed_partition$train, family = binomial(link = "logit")), 
                              direction = "forward", 
                              scope = formula(~ mrace6 + risk_level_cat + meduc + mager9, bfacil3), 
                              trace = FALSE)
# Print the selected model
summary(model_stepwise_AIC)

model_stepwise_AIC
# Test for the effect of race on infant mortality using odds ratios
#exp(cbind(OR = coef(model_stepwise_AIC), confint(model_stepwise_AIC)))
#exp(cbind(OR = coef(model_stepwise_AIC), confint(model_stepwise_AIC, level = .95)))
```

# Manually builing to generate Stepwise plot
```{r}
step1 <- function(response, predictors, candidates, partition) {
  
  # Building the logistic regression
  rhs <- paste0(paste0(predictors, collapse="+"), "+", candidates)
  formulas <- lapply(paste0(response, "~", rhs), as.formula)
  
  # Fit the models and calculate AIC
  aics <- sapply(formulas, function(fm) AIC(glm(fm, data=partition$train, family=binomial)))
  names(aics) <- candidates
  
  # Set the best AIC model as attribute
  attr(aics, "best") <- aics[which.min(aics)]
  
  # Return the AICs
  aics
}

```

# Empty model 
```{r}
model <- NULL
```

Step 1 (no variables):

```{r}
preds <- "1"
cands <- c("mrace6", "risk_level_cat", "meduc", "mager9",
          "bfacil3")
s1 <- step1("mortality_status", preds, cands, imputed_partition)

model <- c(model, attr(s1, "best"))
s1
```

Step 2:

```{r}
preds <- "meduc"
cands <- c("mrace6", "risk_level_cat", "mager9",
          "bfacil3")
s2 <- step1("mortality_status", preds, cands, imputed_partition)

model <- c(model, attr(s2, "best"))
s2
```

Step 3:

```{r}
preds <- c("meduc", "risk_level_cat")
cands <- c("mrace6", "mager9",
          "bfacil3")

s3 <- step1("mortality_status", preds, cands, imputed_partition)

model <- c(model, attr(s3, "best"))
s3
```

Step 4:
```{r}
preds <- c("meduc", "risk_level_cat", "mrace6")
cands <- c("mager9","bfacil3")

s4 <- step1("mortality_status", preds, cands, imputed_partition)

model <- c(model, attr(s4, "best"))
s4
```

Step 5:
```{r}
preds <- c("meduc", "risk_level_cat", "mrace6", "mager9")
cands <- c("bfacil3")

s5 <- step1("mortality_status", preds, cands, imputed_partition)

model <- c(model, attr(s5, "best"))
s5
```

# AIC Plot
```{r}
step_model_aic <- tibble(index=seq_along(model),
                     variable=factor(names(model), levels=names(model)),
                     AIC=model)

ggplot(step_model_aic, aes(y=AIC)) +
  geom_point(aes(x=variable)) +
  geom_line(aes(x=index)) +
  labs(title="Stepwise Model Selection") +
  theme_minimal()
```

# Conclusions 
  # Individuals with lower education levels and younger age may be at a higher risk of mortality.(negative relationship)
  # As risk level and race increases (non-white), the likelihood of mortality_status increases (positive relationship)
  # Variables of significance: meduc (Education), risk_level_cat (Health Risk Level), mrace6 (Race), mager9 (Age)
  
# --------------------------  ANOVA TEST  --------------------------------------

# Original Anova
```{r}
# Fitting the reduced model, excluding mrace6
reduced_model <- glm(mortality_status ~ meduc + mager9 + risk_level_cat, 
                     family = binomial(link = "logit"), 
                     data = imputed_partition$train)
# Fitting the full model with all predictor variables
full_model <- glm(mortality_status ~ meduc + mrace6 + mager9 + risk_level_cat, 
                  family = binomial(link = "logit"), 
                  data = imputed_partition$train)
# Compare the two models using ANOVA
anova(reduced_model, full_model, test = "Chi")
```

# Testing model without age because of stepwise plot)
```{r}
# Fitting the reduced model, excluding mager9
reduced_model_test <- glm(mortality_status ~ meduc + mrace6 + risk_level_cat, 
                     family = binomial(link = "logit"), 
                     data = imputed_partition$train)
# Fitting the full model with all predictor variables
full_model_test <- glm(mortality_status ~ meduc + mrace6 + mager9 + risk_level_cat, 
                  family = binomial(link = "logit"), 
                  data = imputed_partition$train)
# Compare the two models using ANOVA
anova(reduced_model_test, full_model_test, test = "Chi")
```




















<!-- ## PART B -->
<!-- # --------------------------  CLASSIFICIATION/PREDICTION  ---------------------- -->
<!-- # --------------------------  SAMPLING  ---------------------------------------- -->
<!-- # Undersampling: Balancing the Births and Deaths -->
<!-- ```{r} -->
<!-- # Subset the data frame by death and alive classes -->
<!-- training <- imputed_partition$train$data -->
<!-- death_cases <- full_data_clean[training$mortality_status == 1, ] -->
<!-- death_cases -->
<!-- alive_cases <- full_data_clean[training$mortality_status == 0, ] -->
<!-- # Compute the number of cases in the minority class -->
<!-- n_death <- nrow(death_cases) -->
<!-- # Sample n_alive cases from the majority class  -->
<!-- n_alive <- n_death  -->
<!-- training_data <- rbind(death_cases, alive_cases) -->
<!-- ``` -->

<!-- # Undersampling: Balancing the overrepresentation of the Maternal White Population -->
<!-- ```{r} -->
<!-- # Calculate the weights for each race -->
<!-- # weights are based on 2020 demographic percentages (else is based on mean of these percents) -->
<!-- weights <- ifelse(training_data$mrace6 == 1, 1/0.758, ifelse(training_data$mrace6 == 2, 1/0.136, ifelse(training_data$mrace6 == 3, 1/0.013, ifelse(training_data$mrace6 == 4, 1/0.061, ifelse(training_data$mrace6 == 5, 1/0.003, ifelse(training_data$mrace6 == 6, 1/0.029, 1/0.17)))))) -->
<!-- # Balancing the data based on the weights -->
<!-- training_data <- training_data[sample(nrow(training_data), n_alive, replace = FALSE, prob = weights), ]  -->
<!-- training_data <- mutate(training_data, mrace6 = as.factor(mrace6), ilive = as.factor(ilive), meduc = as.factor(meduc), bfacil3 = as.factor(bfacil3), -->
<!--                         no_risks = as.factor(no_risks), mager9 = as.factor(mager9),  -->
<!--                         mortality_status = as.factor(mortality_status), risk_level_cat = as.factor(risk_level_cat)) -->

<!-- # Combine the resampled data frames into a new data frame -->
<!-- head(training_data) -->
<!-- ``` -->

<!--  # WILL DO LOGISTIC REGRESSION WITH TRAINING (OVERSAMPLED) -->
<!--    # TEST SET IS NOT TO BE RESAMPLED -->
<!--    # TRAINING SET WILL BE OVERSAMPLED -->
<!--    # EVALUATION TO BE DONE WITH TEST SET -->
<!--    # CAN USE VARIABLES FROM AIC  -->


<!-- # --------------------------  ACCURACY EVALUATION  ----------------------------- -->



<!-- # --------------------------  SENSITIVITY ANALYSIS ----------------------------- -->
<!-- #STRETCH GOAL WORK  -->

<!-- ```{r} -->
<!-- # fit the logistic regression model -->
<!-- model <- glm(mortality_status ~ mrace6 + mager9 + meduc, data = full_data_clean, family = binomial) -->
<!-- summary(model) -->
<!-- ``` -->


<!-- ```{r} -->
<!-- # may end up needing to weight data  by race after all but luckily there's infant mortality rates documented by race in the documentation maybe we can use those to weight  -->
<!-- # The infant mortality rates based on the documentation for maternal race: -->
<!-- # Overall = 5.51 -->
<!-- # Non-Hispanic White = 4.45 -->
<!-- # Non-Hispanic Black = 10.6 -->
<!-- # Non-Hispanic American Indian or Alaska Native = 7.63 -->
<!-- # Non-Hispanic Asian = 3.25 -->
<!-- # Non-Hispanic Native Hawaiian or Other Pacific Islander = 7.68 -->
<!-- # Hispanic = 4.87 -->
<!-- ``` -->

<!-- ```{r} -->
<!-- data_race <- full_data_clean -->
<!-- data_race$mrace6 <- 1 -->
<!-- data_race -->
<!-- fitted_probs <- predict(model, newdata = data_race, type = "response") -->
<!-- sim_outcome_1 <- rbinom(n = nrow(data_race), size = 1, prob = fitted_probs) -->
<!-- data_race$mrace6 <- 2 -->
<!-- fitted_probs <- predict(model, newdata = data_race, type = "response") -->
<!-- sim_outcome_2 <- rbinom(n = nrow(data_race), size = 1, prob = fitted_probs) -->
<!-- # perform t-test -->
<!-- p_val <- t.test(sim_outcome_2, sim_outcome_1)$p.value -->
<!-- p_val -->
<!-- ``` -->


<!-- ```{r} -->
<!-- # set up the range of values to vary -->
<!-- #race_values <- unique(full_data_clean$mrace6) -->
<!-- race_values <- seq(from = 2, to = 6, by = 1) -->
<!-- # create function to simulate outcome data based on varying maternal race values -->
<!-- # got this function w/ help from chatgpt -->
<!-- sim_outcome <- function(data, maternal_race) { -->
<!--   data$maternal_race <- ifelse(data$mrace6 == maternal_race, 0, 1) -->
<!--   fitted_probs <- predict(model, newdata = data, type = "response") -->
<!--   sim_outcome <- rbinom(n = nrow(data), size = 1, prob = fitted_probs) -->
<!--   return(sim_outcome) -->
<!-- } -->
<!-- # simulate outcome data for varying maternal race values -->
<!-- sim_data <- lapply(race_values, function(mr) sim_outcome(data_race, mr)) -->
<!-- sim_data <- do.call(cbind, sim_data) -->
<!-- # calculate odds ratios for each simulated dataset -->
<!-- odds_ratios <- apply(sim_data, 2, function(x) { -->
<!--   model_sim <- glm(x ~ mrace6 + mager9 + meduc, data = full_data_clean_oversampled, family = binomial) -->
<!--   exp(coef(model_sim)["mrace6"]) -->
<!-- }) -->
<!-- odds_ratios -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # plot odds ratios as a function of maternal race -->
<!-- plot(race_values, odds_ratios, type = "l", xlab = "Maternal Race", ylab = "Odds Ratio") -->
<!-- ``` -->


<!-- ```{r} -->
<!-- p_values <- apply(sim_data, 2, function(x) { -->
<!--   model_sim <- glm(x ~ mrace6 + mager9 + meduc, data = full_data_clean_oversampled, family = binomial) -->
<!--   summary(model_race)$coef[2,4] -->
<!-- }) -->
<!-- p_values -->

<!-- ``` -->

<!-- ```{r} -->
<!-- p_values -->
<!-- # plot p-values for each maternal race group -->
<!-- barplot(p_values, names.arg = race_values, ylab = "p-value") -->
<!-- abline(h = 0.05, lty = 2) -->
<!-- ``` -->



