---
title: "5110 project"
author: "Lesrene Browne"
date: "2023-04-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tibble)
library(dplyr)
library(dtplyr)
library(tidyr)
library(data.table)
library(naniar)
# Reading the data into DFs 
# Numerator File: 
#include deaths to all infants in that same calendar year for which the 
#death certificate can be linked to a birth certificate in the denominator file.
#Graziano path
#numerator_data <- fread("~/Desktop/Project_Intro_Data_Management/link2020usnum.csv", select=c('mrace6'))
#denom_data <- fread("/Users/grazianoperegrino/Desktop/Project_Intro_Data_Management/link2020usden.csv", select=c('mrace6'))
#Denominator File: 
#contains all births occurring in a given calendar year.
# Numerator File
# Tidy Process
  # Read in the data
  # Arrange birthweight in desc order
  # Replace all 9999 values in birthweight and DOB Time with NAs and omit
  # Create unique key from birthweight, DOB time, facility type, pre-pregnancy weight, weight gain 
  # Remove duplicate keys 
# Generating Key
numerator_data <- fread("/Users/lesrene/Downloads/link2020usnum.csv", 
                      select=c('brthwgt','dob_tt', 'bfacil','pwgt_r', 'wtgain')) %>% arrange(desc(brthwgt))
numerator_data <- data.frame(numerator_data)
numerator_data$brthwgt <- replace(numerator_data$brthwgt, numerator_data$brthwgt == 9999, NA)
numerator_data$dob_tt <- replace(numerator_data$dob_tt, numerator_data$dob_tt == 9999, NA)
numerator_data <- numerator_data %>% na.omit()
numerator_data <- numerator_data %>% mutate(key = paste(numerator_data$brthwgt,
                                            numerator_data$dob_tt,
                                            numerator_data$bfacil,
                                            numerator_data$pwgt_r,
                                            numerator_data$wtgain,
                                            sep = '-'))
numerator_data <- distinct(numerator_data)
# Denominator File
denom_data <- fread("/Users/lesrene/Downloads/link2020usden.csv",select=c('brthwgt','dob_tt','bfacil','pwgt_r','wtgain','mrace6','ilive','meduc','bfacil3','rf_pdiab','rf_gdiab','rf_phype','rf_ghype','rf_ehype','rf_ppb','rf_inft','rf_drg','rf_art','rf_cesar','no_risks'))
denom_data <- data.frame(denom_data)
denom_data <- denom_data %>% arrange(desc(brthwgt)) 
denom_data$brthwgt <- replace(denom_data$brthwgt, denom_data$brthwgt == 9999, NA)
denom_data$dob_tt <- replace(denom_data$dob_tt, denom_data$dob_tt == 9999, NA) 
denom_data <- denom_data %>% na.omit()
denom_data <- denom_data %>% mutate(key = paste(denom_data$brthwgt,
                                            denom_data$dob_tt,
                                            denom_data$bfacil,
                                            denom_data$pwgt_r,
                                            denom_data$wtgain,
                                            sep = '-'))
denom_data <- distinct(denom_data)
# Creating the mortality indicator column 
# create a new column in the births dataframe to indicate mortality status
denom_data$mortality_status <- 0
# join the two dataframes based on the primary key column
merged_df <- left_join(denom_data, numerator_data, by = "key")
# update the mortality status column based on whether or not a match was found
merged_df$mortality_status[!is.na(merged_df$brthwgt.y)] <- 1
```

```{r}
library(tibble)
# Removing columns that created the key
births <- denom_data %>% select(-c('brthwgt','dob_tt',
                                                  'bfacil',
                                                 'pwgt_r', 'wtgain'))
# Next steps: 
  # Convert the Y/N in risk columns to 0 and 1's
  # Sum up all the fisk factor columns for risk score
  # Create risk level column and create category (Low, Medium, High) based on score
imputed_rf <- births %>% 
  mutate_at("ilive", ~str_replace(.,"N", "0")) %>%
  mutate_at("ilive", ~str_replace(.,"Y", "1")) %>%
  mutate_at(vars(contains("rf_")), ~str_replace(.,"N", "0")) %>%
  mutate_at(vars(contains("rf_")), ~str_replace(.,"Y", "1")) %>%
  mutate_at(vars(contains("rf_")), ~str_replace(.,"X", "0")) %>%
  mutate_at(vars(contains("rf_")), ~str_replace(.,"U", "0")) %>%
  mutate_at(vars(contains("rf_")), ~as.numeric(as.character(.))) %>%
  mutate(sumrf = rowSums(select(., contains("rf_")))) %>%
  mutate(risk_level = ifelse(sumrf %in% 0:5, "Low", ifelse(sumrf %in% 6:11, "Medium",
                      ifelse(sumrf %in% 12:17, "High", NA))))


tibble(imputed_rf)
```

# Step 1: Calculate the risk level based on columns 313-331
# Step 2: Create 3 categories based on risk level number: no/low risk, medium risk, high risk
Scale: 0 - 10
        Low: Scores from 0 to 5
        Medium: Scores from 6 to 11
        High: Scores from 12 to 17
# Step 3: Use this variable to see if it will make the model better fit 

```{r}
# Visualization 
tibble(numerator_data)
tibble(denom_data)
#numerator_data
numerator_counts <- numerator_data %>%
  select(mrace6) %>%
  count(mrace6) %>%
  rename(num_counts = n) %>%
  group_by(mrace6)
numerator_counts <- data.frame(numerator_counts)
denom_counts <- denom_data %>%
  select(mrace6) %>%
  count(mrace6) %>%
  rename(denom_counts = n) %>%
  group_by(mrace6)
denom_counts <- data.frame(denom_counts)
combined_data <- cbind.data.frame(numerator_counts, denom_counts$denom_counts)%>%
  mutate(proportion = (num_counts/denom_counts$denom_counts)*1000)
combined_data
```



```{r}
library(ggplot2)
ggplot(data, aes(fill=mrace6, y=proportion, x=mrace6)) + 
    geom_bar(stat = "identity")+ theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(title="Race Distribution for Trans People (Weighted & Unweighted)")
```

Does Race affect Infant Mortality Rate ?

- H0: there is no relationship between Race and IMR

- H1: there is a relationship between Race and IMR

```{r}
datatest <- data1 %>%
  #filter(cut %in% c("Fair", "Ideal")) %>%
  transmute(co_seqnum, co_dodyy)
datatest <- na.omit(datatest)
datatest
data2$co_dodyy
diamonds2class
typeof(data1$co_seqnum)
typeof(data2$co_seqnum)
              
g <- inner_join()
g
```